### 2015-07-25 paper/paper.Rmd
added bootstrap cis

```{r load_packages, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
library(plyr)
library(dplyr)
library(stringr)
library(reshape2)
library(ggplot2)
library(lubridate)
#library(ggvis)
library(maps)
library(ggmap)
library(htmltools)
library(rworldmap)
library(grid)    
library(scales)
library(doBy)
library("ggsubplot")
```

```{r load_data, echo = FALSE, message = FALSE, warning = FALSE, results='hide', cache=TRUE}
# This is the scaffolding to import the data
sets <- c("item", "parent", "school", "scoredItem", "student")

@@

world <- getMap(resolution = "low")
#library(plyr)
world.polys <- extractPolygons(world)
detach("package:plyr")
```

```{r dataprep, echo = FALSE, message = FALSE, warning = FALSE}
colnames(student2012)[1] <- "name"
student2012$name <- as.character(student2012$name)
colnames(school2012)[1] <- "name"
school2012$name <- as.character(school2012$name)

@@

```{r gendermath, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
student2012.sub <- student2012[, c(1, 12, 501, 541, 546, 634)]
student2012.sub.summary.gap <- summarise(group_by(student2012.sub, name), 
                  mathgap=mean(PV1MATH[ST04Q01=="Male"], na.rm=T)-
                          mean(PV1MATH[ST04Q01=="Female"], na.rm=T),
                  wmathgap=weighted.mean(PV1MATH[ST04Q01=="Male"], w=SENWGT_STU[ST04Q01=="Male"], na.rm=T)-
                          weighted.mean(PV1MATH[ST04Q01=="Female"], w=SENWGT_STU[ST04Q01=="Female"], na.rm=T),
                  mtest.stat = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$statistic, 
                  mp.value = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$p.value,
                  readgap=mean(PV1READ[ST04Q01=="Male"], na.rm=T)-mean(PV1READ[ST04Q01=="Female"], na.rm=T),
                  rtest.stat = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$statistic, 
                  rp.value = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$p.value,
                  sciencegap=mean(PV1SCIE[ST04Q01=="Male"], na.rm=T)-mean(PV1SCIE[ST04Q01=="Female"], na.rm=T),
                  stest.stat = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$statistic, 
                  sp.value = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$p.value,                                         
                  minmale=min(PV1MATH[ST04Q01=="Male"], na.rm=T), 
                  minfemale=min(PV1MATH[ST04Q01=="Female"], na.rm=T), 
                  maxmale=max(PV1MATH[ST04Q01=="Male"], na.rm=T), 
                  maxfemale=max(PV1MATH[ST04Q01=="Female"], na.rm=T), 
                  propmale=length(PV1MATH[ST04Q01=="Male"])/length(PV1MATH), 
                  propfemale=length(PV1MATH[ST04Q01=="Female"])/length(PV1MATH))
#qplot(mathgap, wmathgap, data=student2012.sub.summary.gap, xlab="Mean", ylab="Weighted Mean",
#      xlim=c(-30,30), ylim=c(-30,30)) + geom_abline(slope=1) + theme(aspect.ratio=1)
student2012.sub.summary.gap$msig <- ifelse(student2012.sub.summary.gap$mp.value>0.05, "none", TRUE)
student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat>0] <- "male"
student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat<0] <- "female"
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$mathgap)])
qplot(name, mathgap, data=student2012.sub.summary.gap, color=msig) + 
  xlab("") +  
  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  scale_y_continuous("Math Score Gap", breaks=seq(-30, 30, 5), limits=c(-30, 30)) + 
   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
```

@@

```{r genderead, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
student2012.sub.summary.gap$rsig <- ifelse(student2012.sub.summary.gap$rp.value>0.05, "none", TRUE)
student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat>0] <- "male"
student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat<0] <- "female"
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$readgap)])
qplot(name, readgap, data=student2012.sub.summary.gap, color=rsig) + 
  xlab("") +
  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  scale_y_continuous("Reading Score Gap") + 
   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
#ggsave("genderreadgap.pdf", width=5, height=8)
#student2012.sub.summary.gap$ssig <- ifelse(student2012.sub.summary.gap$sp.value>0.05, "none", TRUE)
#student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat>0] <- "male"
#student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat<0] <- "female"
#student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
#      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$sciencegap)])
#qplot(name, sciencegap, data=student2012.sub.summary.gap, size=propmale, color=ssig) + 
#  xlab("") +
#  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
#  scale_y_continuous("Science Score Gap") + 
#  scale_size("Prop male") +
#   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw()
```

@@

```{r age, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
student2012.sub <- student2012[, c(1, 6, 12, 14, 501, 541, 546, 634)]
student2012.sub.age <- summarise(group_by(student2012.sub, name), 
                  math4 = mean(PV1MATH[ST06Q01 == 4], na.rm = T), 
                  math5 = mean(PV1MATH[ST06Q01 == 5], na.rm = T), 
                  math6 = mean(PV1MATH[ST06Q01 == 6], na.rm = T), 
                  math7 = mean(PV1MATH[ST06Q01 == 7], na.rm = T), 
                  math8 = mean(PV1MATH[ST06Q01 >  7], na.rm = T))
student2012.sub.age.m <- melt(student2012.sub.age)
student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
student2012.sub.agemedian <- summarise(group_by(student2012.sub, name), medianstart = median(ST06Q01, na.rm=T))
orderedbymedianage <- student2012.sub.agemedian$name[order(student2012.sub.agemedian$medianstart)]
student2012.sub.age.m.median <- left_join(student2012.sub.age.m, student2012.sub.agemedian)
student2012.sub.age.m.median$medianstart <- factor(student2012.sub.age.m.median$medianstart)
student2012.sub.age.m.median$name <- factor(student2012.sub.age.m.median$name, levels=orderedbymedianage)
qplot(variable, value, data=student2012.sub.age.m.median, xlab="Age at start", ylab="Math score", 
      geom=c("point","smooth"), se=F, group=name, colour=medianstart) + 
 facet_wrap(~name, ncol=8) + theme_bw() + theme(legend.position="none")
# ggsave("school-start.pdf", width=14, height=14)
#student2012.sub.age <- summarise(group_by(student2012.sub, name), 
#                  read4 = mean(PV1READ[ST06Q01 == 4], na.rm = T), 
#                  read5 = mean(PV1READ[ST06Q01 == 5], na.rm = T), 
#                  read6 = mean(PV1READ[ST06Q01 == 6], na.rm = T), 
#                  read7 = mean(PV1READ[ST06Q01 == 7], na.rm = T), 
#                  read8 = mean(PV1READ[ST06Q01 >  7], na.rm = T))
#student2012.sub.age.m <- melt(student2012.sub.age)
#student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
#student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
#qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Reading Score", 
#      geom=c("point","smooth"), se=F, group=name) + 
# facet_wrap(~name, ncol=8) + theme_bw() 
#student2012.sub.age <- summarise(group_by(student2012.sub, name), 
#                  scie4 = mean(PV1SCIE[ST06Q01 == 4], na.rm = T), 
#                  scie5 = mean(PV1SCIE[ST06Q01 == 5], na.rm = T), 
#                  scie6 = mean(PV1SCIE[ST06Q01 == 6], na.rm = T), 
#                  scie7 = mean(PV1SCIE[ST06Q01 == 7], na.rm = T), 
#                  scie8 = mean(PV1SCIE[ST06Q01 >  7], na.rm = T))
#student2012.sub.age.m <- melt(student2012.sub.age)
#student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
#student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
#qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Science Score", 
#      geom=c("point","smooth"), se=F, group=name) + 
# facet_wrap(~name, ncol=8) + theme_bw() 
```

```{r agemaps, fig.width=12, fig.height=7, echo = FALSE, message = FALSE, warning = FALSE}
student2012.sub.agemedian$medianstart <- factor(student2012.sub.agemedian$medianstart)
student2012.sub.map <- left_join(student2012.sub.agemedian, world.polys)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=medianstart)) +
  scale_fill_discrete("Median start age") +
  new_theme_empty + theme(legend.position="bottom")
#ggsave("agestartmap.pdf", width=12, height=7)
```

@@

```{r studying, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
student.luke <- student2012[,c("name","PV1MATH","PV1READ","PV1SCIE","ST57Q01")]
#study.luke <- melt(student.luke,id=c("name","ST57Q01"))
studyplot <- 
  ggplot(
    data = student.luke, 
    aes(x = ST57Q01, y = PV1MATH)
  ) + 
  xlab("Hours spent out of school studying per week") + 
  ylab("Math score") + 
  stat_smooth(se=TRUE) +
  facet_wrap(~name)

@@

```{r par_athome, fig.width=12, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, data = athome) + ylab("Math score") + facet_wrap(~name, scales="free_y") + scale_x_discrete("Family structure", labels=c("Mother"="M", "Father"="F", "Both"="B", "Neither"="N")) 
phome
```

@@

```{r parentsjob, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
studentLuke <- student2012
studentLuke$ST15Q01 <- addNA(studentLuke$ST15Q01)
studentLuke$ST19Q01 <- addNA(studentLuke$ST19Q01)

@@

Albania appears to be an outlier in many educational trends.  Many socioeconomic and school factors that play a role in other nations' test scores do not appear to affect Albania's test scores.  For example, Albania has one of the smallest gender gaps for math scores.  Additionally, number of possessions and truancy rates affected test scores for every nation except Albania.

```{r clean, echo=FALSE, message = FALSE, warning = FALSE}
#student2012$CNT = student2012$name
student2012$OECD <- as.character(student2012$OECD)
#student2012[student2012$name == "Albania", ]$OECD = "Albania"
student2012$math = as.numeric(student2012[, "PV1MATH"])
student2012$read = as.numeric(student2012[, "PV1READ"])
student2012$sci = as.numeric(student2012[, "PV1SCIE"])

@@

```{r albania-possessions, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
#df = melt(student2012.sub %>% group_by(name, numposs, OECD) %>% summarise(mmath = mean(math), msci = mean(sci), mread = mean(read)), id=c("name", "numposs", "OECD"))
df = melt(student2012.sub %>% group_by(name, numposs, OECD) %>% summarise(mmath = mean(math)), id=c("name", "numposs", "OECD"))

@@

```{r truancyschool, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
#school2012$name = school2012$name
school2012.sub = school2012 %>% dplyr::select(name, SCHOOLID, SC22Q01)
student2012.sub = student2012 %>% dplyr::select(name, SCHOOLID, math, read, sci, OECD)
#truancy.df = melt( student2012.sub %>% 
#  group_by(name, SCHOOLID) %>% 
#  left_join(school2012.sub, by = c("name", "SCHOOLID")) %>% 
#  group_by(name, SC22Q01, OECD) %>% summarise(math = mean(math), read = mean(read), sci = mean(sci)), 
#  id = c("name", "SC22Q01", "OECD")) %>% filter(!is.na(SC22Q01))
truancy.df = melt( student2012.sub %>% 
  group_by(name, SCHOOLID) %>% 
  left_join(school2012.sub, by = c("name", "SCHOOLID")) %>% 
  group_by(name, SC22Q01, OECD) %>% summarise(math = mean(math)), 
  id = c("name", "SC22Q01", "OECD")) %>% filter(!is.na(SC22Q01))
levels(truancy.df$SC22Q01) = c("Not", "Little", "Some", "Lot")

>>>>

```{r load_packages, cache=FALSE, echo = FALSE, message = FALSE, warning = FALSE, results='hide'}
library(dplyr)
library(plyr)
library(stringr)
library(reshape2)
library(ggplot2)
library(lubridate)
#library(ggvis)
library(maps)
library(ggmap)
library(htmltools)
library(rworldmap)
library(grid)    
library(scales)
library(doBy)
library(boot)
```

```{r load_data, echo = FALSE, message = FALSE, warning = FALSE, results='hide', cache=FALSE}
# This is the scaffolding to import the data
sets <- c("item", "parent", "school", "scoredItem", "student")

@@

world <- getMap(resolution = "low")
#library(plyr)
world.polys <- extractPolygons(world)
#detach("package:plyr")
```

```{r dataprep, cache=FALSE, echo = FALSE, message = FALSE, warning = FALSE}
colnames(student2012)[1] <- "name"
student2012$name <- as.character(student2012$name)
colnames(school2012)[1] <- "name"
school2012$name <- as.character(school2012$name)

@@

```{r gendermath, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
#library(plyr)
student2012.sub <- student2012[, c(1, 12, 501, 541, 546, 634)]
student2012.sub.summary.gap <- summarise(group_by(student2012.sub, name), 
                  mathgap=mean(PV1MATH[ST04Q01=="Male"], na.rm=T)-
                          mean(PV1MATH[ST04Q01=="Female"], na.rm=T),
                  wmathgap=weighted.mean(PV1MATH[ST04Q01=="Male"], w=SENWGT_STU[ST04Q01=="Male"], na.rm=T)-
                          weighted.mean(PV1MATH[ST04Q01=="Female"], w=SENWGT_STU[ST04Q01=="Female"], na.rm=T),
                  mtest.stat = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$statistic, 
                  mp.value = t.test(PV1MATH[ST04Q01=="Male"], PV1MATH[ST04Q01=="Female"])$p.value,
                  readgap=mean(PV1READ[ST04Q01=="Male"], na.rm=T)-mean(PV1READ[ST04Q01=="Female"], na.rm=T),
                  rtest.stat = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$statistic, 
                  rp.value = t.test(PV1READ[ST04Q01=="Male"], PV1READ[ST04Q01=="Female"])$p.value,
                  sciencegap=mean(PV1SCIE[ST04Q01=="Male"], na.rm=T)-mean(PV1SCIE[ST04Q01=="Female"], na.rm=T),
                  stest.stat = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$statistic, 
                  sp.value = t.test(PV1SCIE[ST04Q01=="Male"], PV1SCIE[ST04Q01=="Female"])$p.value,         
                  minmale=min(PV1MATH[ST04Q01=="Male"], na.rm=T), 
                  minfemale=min(PV1MATH[ST04Q01=="Female"], na.rm=T), 
                  maxmale=max(PV1MATH[ST04Q01=="Male"], na.rm=T), 
                  maxfemale=max(PV1MATH[ST04Q01=="Female"], na.rm=T), 
                  propmale=length(PV1MATH[ST04Q01=="Male"])/length(PV1MATH), 
                  propfemale=length(PV1MATH[ST04Q01=="Female"])/length(PV1MATH))
#qplot(mathgap, wmathgap, data=student2012.sub.summary.gap, xlab="Mean", ylab="Weighted Mean",
#      xlim=c(-30,30), ylim=c(-30,30)) + geom_abline(slope=1) + theme(aspect.ratio=1)
student2012.sub.summary.gap$msig <- ifelse(student2012.sub.summary.gap$mp.value>0.05, "none", TRUE)
student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat>0] <- "male"
student2012.sub.summary.gap$msig[student2012.sub.summary.gap$msig==TRUE&student2012.sub.summary.gap$mtest.stat<0] <- "female"
cifn <- function(d, i) {
  x <- d[i,]
  ci <- mean(x$PV1MATH[x$ST04Q01=="Male"], na.rm=T)-
     mean(x$PV1MATH[x$ST04Q01=="Female"], na.rm=T)
  ci
}
bootfn <- function(d) {
  r <- boot(d, statistic=cifn, R=1000)
  l <- sort(r$t)[25]
  u <- sort(r$t)[975]
  ci <- c(l, u)
  return(ci)
}
#ddply(student2012.sub, .(name), bootfn)
#results <- boot(data=aus, statistic=cifn, R=1000)
student2012.sub.summary.gap.boot <- ddply(student2012.sub, .(name), bootfn)
colnames(student2012.sub.summary.gap.boot)[2:3] <- c("ml", "mu")
student2012.sub.summary.gap <- merge(student2012.sub.summary.gap, student2012.sub.summary.gap.boot)
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$mathgap)])
qplot(name, mathgap, data=student2012.sub.summary.gap, color=msig, size=I(3)) + 
  geom_segment(aes(x=name, xend=name, y=ml, yend=mu)) +
  xlab("") +  
  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  scale_y_continuous("Math Score Gap", breaks=seq(-45, 45, 5), limits=c(-45, 45)) + 
   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
```

@@

```{r genderead, fig.width=8, fig.height=12, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE}
student2012.sub.summary.gap$rsig <- ifelse(student2012.sub.summary.gap$rp.value>0.05, "none", TRUE)
student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat>0] <- "male"
student2012.sub.summary.gap$rsig[student2012.sub.summary.gap$rsig==TRUE&student2012.sub.summary.gap$rtest.stat<0] <- "female"
cifn <- function(d, i) {
  x <- d[i,]
  ci <- mean(x$PV1READ[x$ST04Q01=="Male"], na.rm=T)-
     mean(x$PV1READ[x$ST04Q01=="Female"], na.rm=T)
  ci
}
bootfn <- function(d) {
  r <- boot(d, statistic=cifn, R=1000)
  l <- sort(r$t)[25]
  u <- sort(r$t)[975]
  ci <- c(l, u)
  return(ci)
}
#ddply(student2012.sub, .(name), bootfn)
#results <- boot(data=aus, statistic=cifn, R=1000)
student2012.sub.summary.gap.boot <- ddply(student2012.sub, .(name), bootfn)
colnames(student2012.sub.summary.gap.boot)[2:3] <- c("rl", "ru")
student2012.sub.summary.gap <- merge(student2012.sub.summary.gap, student2012.sub.summary.gap.boot)
student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$readgap)])
qplot(name, readgap, data=student2012.sub.summary.gap, color=rsig, size=I(3)) + 
  geom_segment(aes(x=name, xend=name, y=rl, yend=ru)) +
  xlab("") +
  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
  scale_y_continuous("Reading Score Gap") + 
   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw() + theme(legend.position="bottom")
#ggsave("genderreadgap.pdf", width=5, height=8)
#student2012.sub.summary.gap$ssig <- ifelse(student2012.sub.summary.gap$sp.value>0.05, "none", TRUE)
#student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat>0] <- "male"
#student2012.sub.summary.gap$ssig[student2012.sub.summary.gap$ssig==TRUE&student2012.sub.summary.gap$stest.stat<0] <- "female"
#student2012.sub.summary.gap$name <- factor(student2012.sub.summary.gap$name, 
#      levels=student2012.sub.summary.gap$name[order(student2012.sub.summary.gap$sciencegap)])
#qplot(name, sciencegap, data=student2012.sub.summary.gap, size=propmale, color=ssig) + 
#  xlab("") +
#  scale_colour_manual("Significant", values=c("male"="skyblue", "female"="pink", "none"="lightgreen")) +
#  scale_y_continuous("Science Score Gap") + 
#  scale_size("Prop male") +
#   geom_hline(yintercept=0, colour="grey80") + coord_flip() + theme_bw()
```

@@

```{r age, cache=FALSE, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
#library(dplyr)
detach("package:plyr")
student2012.sub <- student2012[, c(1, 6, 12, 14, 501, 541, 546, 634)]
student2012.sub.age <- summarise(group_by(student2012.sub, name), 
                  math4 = mean(PV1MATH[ST06Q01 == 4], na.rm = T), 
                  math5 = mean(PV1MATH[ST06Q01 == 5], na.rm = T), 
                  math6 = mean(PV1MATH[ST06Q01 == 6], na.rm = T), 
                  math7 = mean(PV1MATH[ST06Q01 == 7], na.rm = T), 
                  math8 = mean(PV1MATH[ST06Q01 >  7], na.rm = T))
student2012.sub.age.m <- melt(student2012.sub.age)
student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
student2012.sub.agemedian <- summarise(group_by(student2012.sub, name), medianstart = median(ST06Q01, na.rm=T))
orderedbymedianage <- student2012.sub.agemedian$name[order(student2012.sub.agemedian$medianstart)]
student2012.sub.age.m.median <- left_join(student2012.sub.age.m, student2012.sub.agemedian, by="name")
student2012.sub.age.m.median$medianstart <- factor(student2012.sub.age.m.median$medianstart)
student2012.sub.age.m.median$name <- factor(student2012.sub.age.m.median$name, levels=orderedbymedianage)
qplot(variable, value, data=student2012.sub.age.m.median, xlab="Age at start", ylab="Math score", 
      geom=c("point","smooth"), se=F, group=name, colour=medianstart) + 
 facet_wrap(~name, ncol=8) + theme_bw() + theme(legend.position="none")
# ggsave("school-start.pdf", width=14, height=14)
#student2012.sub.age <- summarise(group_by(student2012.sub, name), 
#                  read4 = mean(PV1READ[ST06Q01 == 4], na.rm = T), 
#                  read5 = mean(PV1READ[ST06Q01 == 5], na.rm = T), 
#                  read6 = mean(PV1READ[ST06Q01 == 6], na.rm = T), 
#                  read7 = mean(PV1READ[ST06Q01 == 7], na.rm = T), 
#                  read8 = mean(PV1READ[ST06Q01 >  7], na.rm = T))
#student2012.sub.age.m <- melt(student2012.sub.age)
#student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
#student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
#qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Reading Score", 
#      geom=c("point","smooth"), se=F, group=name) + 
# facet_wrap(~name, ncol=8) + theme_bw() 
#student2012.sub.age <- summarise(group_by(student2012.sub, name), 
#                  scie4 = mean(PV1SCIE[ST06Q01 == 4], na.rm = T), 
#                  scie5 = mean(PV1SCIE[ST06Q01 == 5], na.rm = T), 
#                  scie6 = mean(PV1SCIE[ST06Q01 == 6], na.rm = T), 
#                  scie7 = mean(PV1SCIE[ST06Q01 == 7], na.rm = T), 
#                  scie8 = mean(PV1SCIE[ST06Q01 >  7], na.rm = T))
#student2012.sub.age.m <- melt(student2012.sub.age)
#student2012.sub.age.m$variable <- substr(student2012.sub.age.m$variable, 5, 5)
#student2012.sub.age.m$variable <- as.numeric(student2012.sub.age.m$variable)
#qplot(variable, value, data=student2012.sub.age.m, xlab="Age at start", ylab="Science Score", 
#      geom=c("point","smooth"), se=F, group=name) + 
# facet_wrap(~name, ncol=8) + theme_bw() 
```

```{r agemaps, cache=FALSE, fig.width=12, fig.height=7, echo = FALSE, message = FALSE, warning = FALSE}
student2012.sub.agemedian$medianstart <- factor(student2012.sub.agemedian$medianstart)
student2012.sub.map <- left_join(student2012.sub.agemedian, world.polys)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=medianstart)) +
  scale_fill_discrete("Median start age") +
  new_theme_empty + theme(legend.position="bottom")
#ggsave("agestartmap.pdf", width=12, height=7)
```

@@

```{r studying, cache=FALSE, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
student.luke <- student2012[,c("name","PV1MATH","PV1READ","PV1SCIE","ST57Q01")]
#study.luke <- melt(student.luke,id=c("name","ST57Q01"))
studyplot <- 
  ggplot(
    data = student.luke, 
    aes(x = ST57Q01, y = PV1MATH)
  ) + 
  xlab("Hours spent out of school studying per week") + 
  ylab("Math score") + 
  stat_smooth(se=TRUE) +
  facet_wrap(~name)

@@

```{r par_athome, cache=FALSE, fig.width=12, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, data = athome) + ylab("Math score") + facet_wrap(~name, scales="free_y") + scale_x_discrete("Family structure", labels=c("Mother"="M", "Father"="F", "Both"="B", "Neither"="N")) 
phome
```

@@

```{r parentsjob, cache=FALSE, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
studentLuke <- student2012
studentLuke$ST15Q01 <- addNA(studentLuke$ST15Q01)
studentLuke$ST19Q01 <- addNA(studentLuke$ST19Q01)

@@

Albania appears to be an outlier in many educational trends.  Many socioeconomic and school factors that play a role in other nations' test scores do not appear to affect Albania's test scores.  For example, Albania has one of the smallest gender gaps for math scores.  Additionally, number of possessions and truancy rates affected test scores for every nation except Albania.  This most likely indicates arrors in the data collected for the country.

```{r clean, cache=FALSE, echo=FALSE, message = FALSE, warning = FALSE}
#student2012$CNT = student2012$name
student2012$OECD <- as.character(student2012$OECD)
#student2012[student2012$name == "Albania", ]$OECD = "Albania"
student2012$math = as.numeric(student2012[, "PV1MATH"])
student2012$read = as.numeric(student2012[, "PV1READ"])
student2012$sci = as.numeric(student2012[, "PV1SCIE"])

@@

```{r albania-possessions, cache=FALSE, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
#df = melt(student2012.sub %>% group_by(name, numposs, OECD) %>% summarise(mmath = mean(math), msci = mean(sci), mread = mean(read)), id=c("name", "numposs", "OECD"))
df = melt(student2012.sub %>% group_by(name, numposs, OECD) %>% summarise(mmath = mean(math)), id=c("name", "numposs", "OECD"))

@@

```{r truancyschool, cache=FALSE, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
#school2012$name = school2012$name
school2012.sub = school2012 %>% dplyr::select(name, SCHOOLID, SC22Q01)
student2012.sub = student2012 %>% dplyr::select(name, SCHOOLID, math, read, sci, OECD)
#truancy.df = melt( student2012.sub %>% 
#  group_by(name, SCHOOLID) %>% 
#  left_join(school2012.sub, by = c("name", "SCHOOLID")) %>% 
#  group_by(name, SC22Q01, OECD) %>% summarise(math = mean(math), read = mean(read), sci = mean(sci)), 
#  id = c("name", "SC22Q01", "OECD")) %>% filter(!is.na(SC22Q01))
truancy.df = melt( student2012.sub %>% 
  group_by(name, SCHOOLID) %>% 
  left_join(school2012.sub, by = c("name", "SCHOOLID")) %>% 
  group_by(name, SC22Q01, OECD) %>% summarise(math = mean(math)), 
  id = c("name", "SC22Q01", "OECD")) %>% filter(!is.na(SC22Q01))
levels(truancy.df$SC22Q01) = c("Not", "Little", "Some", "Lot")
